---
title: "R project"
author: "Goodness Rex Nze-Igwe"
date: "2024-12-11"
output: html_document
---

## Data Cleaning and Restructuring

### Load the Dataset

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(caret)
```

```{r}
path <- "~/Downloads/AB_NYC_2019.csv"
```

```{r}
# Load the dataset
data <- read.csv(path)

# View the structure of the data
str(data)


```

### Handle Missing Values

```{r}
# Check for missing values
colSums(is.na(data))

# Impute missing values for `reviews_per_month` with 0
data$reviews_per_month[is.na(data$reviews_per_month)] <- 0

# Drop rows with missing `name` or `host_name`
data <- data %>% drop_na(name, host_name)

# Verify no missing values remain
colSums(is.na(data))
```

### Remove Duplicates and Outliers

```{r}
# Remove duplicates
data <- data %>% distinct()

# Remove outliers in the `price` column
data <- data %>% filter(price > 0 & price < quantile(price, 0.99))

# Summary statistics after cleaning
summary(data)
```

## Data Restructuring

### Transforming Variables

```{r}
# Create a new column for price category
data <- data %>% mutate(price_category = case_when(
  price < 60~ "Low",
  price < 120 ~ "Medium",
  TRUE ~ "High"
))



# View the updated structure
glimpse(data)
```

## Data Visualization

### Visualize Price Distribution

```{r}
ggplot(data, aes(x = price)) +
  geom_histogram(binwidth = 60, fill = "green", color = "black") +
  labs(title = "Distribution of Prices", x = "Price", y = "Count") +
  theme_minimal()
```

### Room Type by Neighborhood Group

```{r}
# Calculate average price by neighborhood
avg_price <- data %>% 
  group_by(neighbourhood) %>% 
  summarize(avg_price = mean(price))


ggplot(data, aes(x = neighbourhood_group, fill = room_type)) +
  geom_bar(position = "dodge") +
  labs(title = "Room Types by Neighborhood Group", x = "Neighborhood Group", y = "Count", fill = "Room Type") +
  theme_minimal()
```

### Average Price by Neighborhood

```{r}
avg_price <- data %>% 
  group_by(neighbourhood) %>% 
  summarize(avg_price = mean(price)) %>% 
  mutate(category = if_else(rank(-avg_price) <= 8, neighbourhood, "Other"))

ggplot(avg_price, aes(x = reorder(category, avg_price), y = avg_price)) +
  geom_bar(stat = "identity", fill = "purple") +
  coord_flip() +
  labs(title = "Average Price by Neighborhood (Top 8+ Other)", x = "Neighborhood", y = "Average Price") +
  theme_minimal()

```

## Prediction Model

### Prepare Data for Machine Learning

```{r}
library(caret)

# Select relevant columns and encode categorical variables
data_ml <- data %>% select(price, minimum_nights, number_of_reviews, reviews_per_month, calculated_host_listings_count, availability_365, neighbourhood_group, room_type)
data_ml <- data_ml %>% mutate(
  neighbourhood_group = as.factor(neighbourhood_group),
  room_type = as.factor(room_type)
)

# Split the data into training and testing sets
set.seed(123)
train_index <- createDataPartition(data_ml$price, p = 0.8, list = FALSE)
train_data <- data_ml[train_index, ]
test_data <- data_ml[-train_index, ]
```

### Train a Linear Regression Model

```{r}
train_data <- na.omit(train_data)  ## making sure there is no missing values in the trained data

# Train the model
model <- train(price ~ ., data = train_data, method = "lm")

# Summary of the model
summary(model)

```

### Evaluate the Model

```{r}
# Predict on test data
predictions <- predict(model, newdata = test_data)

# Calculate the Root Mean Squared Error
rmse <- sqrt(mean((test_data$price - predictions)^2))

# Calculate the Mean Absolute Error
mae <- mean(abs(test_data$price - predictions))

# Calculate the Mean Absolute Percentage Error
mape <- mean(abs((test_data$price - predictions) / test_data$price)) * 100

# Print results
list(RMSE = rmse, MAE = mae, MAPE = mape)

```

## Conclusion

This analysis focused on cleaning and restructuring the Airbnb_NYC_2019 dataset, visualizing key patterns with ggplot2, and developing a machine learning model to predict prices. Future improvements could involve experimenting with more sophisticated models or incorporating additional features.

